The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-08 Papers/Universalis/ModalCollapse.tex
Add Square Image

Nevertheless, G{\"o}del's work
still leaves room for criticism. In particular, his axioms are so
strong that they entail a \emph{modal collapse} 
\cite{ToDo:sobel1987?,sobel2004logic}: 
everything that is the case is so necessarily. 
There has been an impressive body of recent and ongoing
work (cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,ToDo:FrodeAndHajek,ContemporaryBibliography} and the references therein)
proposing solutions for the modal collapse. 
The goal of this short note is merely to remark that the modal collapse is a collapse of the modal square of opposition.

@@

A crucial step of most ontological arguments is the claim that 
if God's existence is possible, then it is necessary. 
This is Lemma L2 in G\"odel's proof. 
In the modal square of opposition (Fig. \ref{ToDo}), 
this is an unusual situation in which the \textbf{I} corner 
must imply and entail the \textbf{A} corner, 
in the particular case when $\phi$ is $\exq x G(x)$. 
G\"odel's proof shows that his axioms are strong enough 
to invert the direction of entailment for the sentence at issue. 
The question, however, is whether the axioms are not too strong, 
also allowing the inverted entailment for arbitrary $\phi$. 
That is essentially the question asked by Sobel \cite{Sobel1987}; 
and his proof of the modal collapse (MC) provides an affirmative 
answer. It is possible to show that this form of the modal 
collapse entails (in modal logic K) a collapse of the modal 
square (MC'): the subcontraries entail (and even imply) the 
contraries. Furthermore, in any modal logic where where the 
axiom T holds (i.e. the accessibility relation is reflexive), 
a total collapse of the modalities (MC'') occurs.

@@

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[T] Everything that is necessary is the case:
  $$\hol{\allq \phi [\nec \phi \imp \phi]}$$ 
\item[EI] (Modal Existential Import):
  $$\hol{\pos \top}$$ 
\item[AI] Everything that is necessary is possible:
  $$\hol{\allq \phi [\nec \phi \imp \pos \phi]}$$ 
\item[MC''] Modalities collapse completely:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}

>>>>

Nevertheless, G{\"o}del's work
still leaves room for criticism. In particular, his axioms are so
strong that they entail a \emph{modal collapse} 
\cite{ToDo:sobel1987?,sobel2004logic}: 
everything that is the case is so necessarily. 
There has been an impressive body of recent and ongoing
work (cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,ToDo:FrodeAndHajek,ContemporaryBibliography} and the references therein)
proposing solutions for the modal collapse. 
The goal of this short note is to discuss the modal collapse from the point of view of the modal square of opposition.

@@

A crucial step of most ontological arguments is the claim that 
if God's existence is possible, then it is necessary. 
This is Lemma L2 in G\"odel's proof. 
In the modal square of opposition (Fig. \ref{fig:square}), 
this is an unusual situation in which the \textbf{I} corner 
must imply and entail the \textbf{A} corner, 
in the particular case when $\phi$ is $\exq x G(x)$. 
G\"odel's proof shows that his axioms are strong enough 
to invert the direction of entailment for the sentence at issue. 
The question, however, is whether the axioms are not too strong, 
also allowing the inverted entailment for arbitrary $\phi$. 
That is essentially the question asked by Sobel \cite{Sobel1987}; 
and his proof of the modal collapse (MC) provides an affirmative 
answer. It is possible to show that this form of the modal 
collapse entails (in modal logic K) a collapse of the modal 
square (MC'), causing the subcontraries to entail (and even imply) their respective contraries. Normally, as shown in Fig. \ref{fig:square} in the modal square of opposition, only the other direction of entailment holds: the contraries entail their subcontraries, assuming the \emph{modal existential import} ExImp.

Furthermore, in any modal logic where the 
axiom T holds (i.e. where the accessibility relation is reflexive), a total collapse of the modalities (MC'') occurs. In particular, under this stronger form of modal collapse, the contraries entail their subcontraries even without the existential import.

\begin{figure}[t]
\includegraphics[width=0.9\textwidth]{./SquarePics/ModalSquare.png}
\caption{Modal Square of Opposition.}
\label{fig:square}
\end{figure}

@@

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[T] Everything that is necessary is the case:
  $$\hol{\allq \phi [\nec \phi \imp \phi]}$$ 
\item[ExImp] (Modal Existential Import):
  $$\hol{\pos \top}$$ 
\item[AI] Everything that is necessary is possible:
  $$\hol{\allq \phi [\nec \phi \imp \pos \phi]}$$ 
\item[MC''] Modalities collapse completely:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}

\begin{Huge}

$$
\nec \ \phi
$$

$$
\pos \ \phi
$$

$$
\neg \ \nec \ \phi
$$

$$
\neg \ \pos \ \phi
$$

$$
\pos \ \exists x \ G(x)
$$

$$
\nec \ \exists x \ G(x)
$$

\end{Huge}
